For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/paper.tex
stuff

\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by the National Research,
Development and Innovation Fund of Hungary, financed under the
Thematic Excellence Programme funding scheme, Project
no. ED18-1-2019-0030 (Application-specific highly reliable IT
solutions), by the New National Excellence Program of the Ministry
for Innovation and Technology, Project no. ÚNKP-19-4-ELTE-874, and
by the Bolyai Fellowship of the Hungarian Academy of Sciences,
Project no. BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
\author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@inria.fr}{https://orcid.org/0000-0002-9299-641X}{}

@@

\acknowledgements{The authors would like to thank Thorsten Altenkirch, Simon Boulier, Fredrik Nordvall-Forsberg and Jakob von Raumer for discussions on the topics of this paper.}%optional

@@

\begin{abstract}
  Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work by Altenkirch, Kaposi, and Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda. TODO: rewrite
\end{abstract}

@@

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimsed version
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
hence the two sorts have to be defined simultaneously. Moreover, $\Ty$ is
indexed over $\Con$. This precludes a reduction analogous to the reduction of
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
itself. Another unique feature of IITs (which also shows up in higher inductive
types \cite{HoTTbook}) is that later constructors can refer to previous
constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

The elimination principle for the above IIT has the following two
motives (one for each sort) and four methods (one for each
constructor).
\begin{alignat*}{5}
  & Con^D && : \Con\ra\Set \\
  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & {empty}^D && : Con^D\,\mathsf{empty} \\
  & {ext}^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,({ext}^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
Above we used implicit quantifications for $\Gamma:\Con$ and
$A:\Ty\,\Gamma$ to ease readability, e.g.\ $Ty^D$ has an implicit
parameter $\Gamma$ before its explicit parameter of type
$Con^D\,\Gamma$.

@@

\section{Constructing IITs from the Theory of IIT Signatures}
\label{sec:constructingiits}

@@

\begin{theorem}
  If a model of type theory supports the theory of IIT signatures,
  then it supports all IITs.
\end{theorem}
\begin{proof}
  For a signature $({sig},{sigc})$, we define constructors as
  \[
    \con_{sig}:= (\ll{sig}\,\I\rr_{\IC_{{sig}\,\I}})^\C\,\id^\I : ({sig}\,\ADS)^\A
  \]
  This typechecks as
  $\ll{sig}\,\I\rr_\A = \ll\blank\rr_\A\,({sig}\,\I)
  \overset{{sigc}}{=} {{sig}\,\A} = ({sig}\,\ADS)^\A$. We define the
  eliminator by
  and an
  \[
    \elim_{sig}\,\gamma^D := (\ll{sig}\,\I\rr_{\IE_{\gamma^D}})^\E\,\id^\I  : ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
  This typechecks firstly because the type of $\gamma^D$ matches the
  type of the parameter of $\IE$:
  \[
    ({sig}\,\ADS)^\D\,\con_{sig} \overset{{sigc}}{=}
    (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\D\,\con_{sig} =
    (\ll{sig}\,\I\rr_{\ADS})^\D\,\con_{sig},
  \]
  and the result also has the correct type:
  \[
    \ll{sig}\,\I\rr^\S\,\con_{sig}\,\gamma^D = (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\S\,\con_{sig}\,\gamma^D \overset{{sigc}}{=} ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{proof}

@@

>>>>

\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{this author was supported by the National Research,
Development and Innovation Fund of Hungary, financed under the
Thematic Excellence Programme funding scheme, Project
no.\ ED18-1-2019-0030 (Application-specific highly reliable IT
solutions), by the New National Excellence Program of the Ministry
for Innovation and Technology, Project no.\ ÚNKP-19-4-ELTE-874, and
by the Bolyai Fellowship of the Hungarian Academy of Sciences,
Project no.\ BO/00659/19/3.}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
\author{Andr{\'a}s Kov{\'a}cs}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{this author was supported by the European Union, co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}
\author{Ambroise Lafont}{IMT Atlantique, Inria, LS2N CNRS, Nantes, France}{ambroise.lafont@inria.fr}{https://orcid.org/0000-0002-9299-641X}{}

@@

\acknowledgements{The authors would like to thank Thorsten Altenkirch, Rafa{\"e}l Bocquet, Simon Boulier, Fredrik Nordvall-Forsberg and Jakob von Raumer for discussions on the topics of this paper. We also thank the anonymous reviewers for their helpful comments and suggestions.}%optional

@@

\begin{abstract}
  TODO: rewrite. Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work by Altenkirch, Kaposi, and Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda. TODO: rewrite
\end{abstract}

@@

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimised version
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
hence the two sorts have to be defined simultaneously. Moreover, $\Ty$ is
indexed over $\Con$. This precludes a reduction analogous to the reduction of
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
itself. Another unique feature of IITs (which also holds for higher inductive
types \cite{HoTTbook}) is that later constructors can refer to previous
constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

The elimination principle for the above IIT has the following two
motives (one for each sort) and four methods (one for each
constructor).
\begin{alignat*}{5}
  & Con^D && : \Con\ra\Set \\
  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & {empty}^D && : Con^D\,\mathsf{empty} \\
  & {ext}^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,({ext}^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*} xxx
Above we used implicit quantifications for $\Gamma:\Con$ and
$A:\Ty\,\Gamma$ to ease readability, e.g.\ $Ty^D$ has an implicit
parameter $\Gamma$ before its explicit parameter of type
$Con^D\,\Gamma$.

@@

\section{Constructing all IITs from the Theory of IIT Signatures}
\label{sec:constructingiits}

@@

\begin{theorem}\label{th:tosToIITs}
  If a model of type theory supports the theory of IIT signatures,
  then it supports all IITs.
\end{theorem}
\begin{proof}
  For a signature $({sig},{sigc})$, we define constructors as
  \[
    \con_{sig}:= (\ll{sig}\,\I\rr_{\IC_{{sig}\,\I}})^\C\,\id^\I : ({sig}\,\ADS)^\A
  \]
  This typechecks as
  $\ll{sig}\,\I\rr_\A = \ll\blank\rr_\A\,({sig}\,\I)
  \overset{{sigc}}{=} {{sig}\,\A} = ({sig}\,\ADS)^\A$. We define the
  eliminator by
  and an
  \[
    \elim_{sig}\,\gamma^D := (\ll{sig}\,\I\rr_{\IE_{\gamma^D}})^\E\,\id^\I  : ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
  This typechecks firstly because the type of $\gamma^D$ matches the
  type of the parameter of $\IE$:
  \[
    ({sig}\,\ADS)^\D\,\con_{sig} \overset{{sigc}}{=}
    (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\D\,\con_{sig} =
    (\ll{sig}\,\I\rr_{\ADS})^\D\,\con_{sig},
  \]
  and the result also has the correct type:
  \[
    \ll{sig}\,\I\rr^\S\,\con_{sig}\,\gamma^D = (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\S\,\con_{sig}\,\gamma^D \overset{{sigc}}{=} ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{proof}

@@

In this section we show that any model of ETT which supports indexed
W-types also supports the theory of signatures, and as a consequence
of \Cref{th:tosToIITs}, all IITs. For this, we work in the internal
language of a model of ETT supporting indexed W-types
\cite{indexedcont}. Indexed W-types correspond to the usual notion of
(possibly mutual) indexed inductive types. We will use Agda-style
notation for defining such inductive families: we list the sorts and
constructors and use pattern matching when eliminating from them. qFor
an encoding of mutual inductive families as families as indexed
W-types, see e.g.\ \cite{mutual}.
